dispRity manual

### 2020-11-02 inst/gitbook/08_palaeo-demo.Rmd
bug fix in paleo demo

```{r}
## Creating the vector of time bins ages
(time_bins <- rev(seq(from = 0, to = 100, by = 20)))

## Splitting the morphospace using the chrono.subsets function
(binned_morphospace <- chrono.subsets(data = my_matrix, tree = my_tree,
    method = "discrete", time = time_bins, inc.nodes = FALSE,
    FADLAD = my_fadlad))
```

@@

```{r}
## Bootstrapping each time subset 100 times (default)
(boot_bin_morphospace <- boot.matrix(binned_morphospace))

## Getting the minimum number of rows (i.e. taxa) in the time subsets
min(size.subsets(boot_bin_morphospace))

## Bootstrapping each time subset 100 times and rarefying them 
(rare_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100,
    rarefaction = 6))
```

@@

```{r}
my_test <- test.metric(my_matrix, metric = c(sum, variances), shifts = c("random", "size"))
summary(my_test)
plot(my_test)
```

@@

```{r}
## Calculating disparity for the bootstrapped data
(boot_disparity <- dispRity(boot_bin_morphospace, metric = c(sum, variances)))

## Calculating disparity for the rarefied data
(rare_disparity <- dispRity(rare_bin_morphospace, metric = c(sum, variances)))
```

@@

```{r}
## Summarising the disparity results
summary(boot_disparity)
summary(rare_disparity)
```

@@

## Plotting the bootstrapped and rarefied results
plot(boot_disparity, type = "continuous", main = "bootstrapped results")
plot(rare_disparity, type = "continuous", main = "rarefied results")
```

@@

```{r}
## Testing the differences between bins in the bootstrapped dataset.
test.dispRity(boot_disparity, test = wilcox.test, comparison = "sequential",
    correction = "bonferroni")

## Testing the differences between bins in the rarefied dataset.
test.dispRity(rare_disparity, test = wilcox.test, comparison = "sequential",
    correction = "bonferroni")
```

>>>>

```{r}
## Creating the vector of time bins ages
time_bins <- rev(seq(from = 0, to = 100, by = 20))

## Splitting the morphospace using the chrono.subsets function
binned_morphospace <- chrono.subsets(data = my_matrix, tree = my_tree,
    method = "discrete", time = time_bins, inc.nodes = FALSE,
    FADLAD = my_fadlad)
```

@@

```{r}
## Getting the minimum number of rows (i.e. taxa) in the time subsets
minimum_size <- min(size.subsets(binned_morphospace))

## Bootstrapping each time subset 100 times and rarefying them 
rare_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100,
    rarefaction = minimum_size)
```

@@

```{r, fig.height = 16, fig.width = 8}
my_test <- test.metric(my_matrix, metric = c(sum, variances), shifts = c("random", "size"))
summary(my_test)
plot(my_test)
```

@@

```{r}
## Calculating disparity for the bootstrapped and rarefied data
disparity <- dispRity(rare_bin_morphospace , metric = c(sum, variances))
```

@@

```{r}
## Summarising the disparity results
summary(disparity)
```

@@

## Plotting the bootstrapped and rarefied results
plot(disparity, type = "continuous", main = "bootstrapped results")
plot(disparity, type = "continuous", main = "rarefied results",
     rarefaciton = minimum_size)
```

@@

```{r}
## Testing the differences between bins in the bootstrapped dataset.
test.dispRity(disparity, test = wilcox.test, comparison = "sequential",
    correction = "bonferroni")

## Testing the differences between bins in the rarefied dataset.
test.dispRity(disparity, test = wilcox.test, comparison = "sequential",
    correction = "bonferroni", rarefaction  = minimum_size)
```

### 2020-11-02 inst/gitbook/dispRity_manual.Rmd
bug fix in paleo demo

```{r, eval = TRUE, echo = FALSE}
## master version
version_raw <- RCurl::getURL("https://raw.githubusercontent.com/TGuillerme/dispRity/master/DESCRIPTION")
version_master <- strsplit(strsplit(version_raw, split = "Version: ")[[1]][2], split = "\nDate:")[[1]][1]
version_master <- 1.4

## release version
version_raw <- RCurl::getURL("https://raw.githubusercontent.com/TGuillerme/dispRity/release/DESCRIPTION")
version_release <- strsplit(strsplit(version_raw, split = "Version: ")[[1]][2], split = "\nDate:")[[1]][1]
version_release <- 1.4

## CRAN version
version_raw <- RCurl::getURL("https://cran.r-project.org/web/packages/dispRity/index.html")
version_CRAN <- strsplit(strsplit(version_raw, split = "\n<td>Version:</td>\n<td>")[[1]][2], split = "</td>\n</tr>\n<tr>\n<td>Depends:")[[1]][1]
version_CRAN <- 1.4
```

@@

@@

@@

Level | Name | Description | Source | 
------|------|---------------------------------------------------|--------|
2 | `ancestral.dist` | The distance between an element and its ancestor | `dispRity` |
2 | `angles` | The angle of main variation of each dimensions | `dispRity` |
2 | `centroids`<sup>1</sup> | The distance between each element and the centroid of the ordinated space | `dispRity` |
1 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln$area` |
1 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln$vol` |
2 | `deviations` | The minimal distance between each element and a hyperplane | `dispRity` |
1 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
2 | `displacements` | The ratio between the distance from a reference and the distance from the centroid | `dispRity` |
1 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | Donohue _et al._ (2013) |
1 | `func.div` | The functional divergence (the ratio of deviation from the centroid) | `dispRity` (similar to [`FD`](https://cran.r-project.org/web/packages/FD/index.html)`::dbFD$FDiv` but without abundance)|
1 | `func.eve` | The functional evenness (the minimal spanning tree distances evenness) | `dispRity` (similar to [`FD`](https://cran.r-project.org/web/packages/FD/index.html)`::dbFD$FEve` but without abundance)|
1 | `mode.val` | The modal value | `dispRity` |
1 | `n.ball.volume` | The hyper-spherical (*n*-ball) volume | `dispRity` |
2 | `neighbours` | The distance to specific neighbours (e.g. the nearest neighbours - by default) | `dispRity` |
2 | `pairwise.dist` | The pairwise distances between elements | [`vegan`](https://cran.r-project.org/web/packages/vegan/index.html)`::vegist` |
2 | `quantiles` | The *n*th quantile range per axis | `dispRity` |
2 | `radius` | The radius of each dimensions | `dispRity` |
2 | `ranges` | The range of each dimension | `dispRity` |
2 | `span.tree.length` | The minimal spanning tree length | [`ape`](https://cran.r-project.org/web/packages/vegan/index.html) (but see also [`vegan`](https://cran.r-project.org/web/packages/vegan/index.html)`::spantree`) |
2 | `variances` | The variance of each dimension | `dispRity` |

@@

>>>>

```{r, eval = TRUE, echo = FALSE}
## master version
version_raw <- RCurl::getURL("https://raw.githubusercontent.com/TGuillerme/dispRity/master/DESCRIPTION")
version_master <- strsplit(strsplit(version_raw, split = "Version: ")[[1]][2], split = "\nDate:")[[1]][1]
version_master <- 1.5

## release version
version_raw <- RCurl::getURL("https://raw.githubusercontent.com/TGuillerme/dispRity/release/DESCRIPTION")
version_release <- strsplit(strsplit(version_raw, split = "Version: ")[[1]][2], split = "\nDate:")[[1]][1]
version_release <- 1.5

## CRAN version
version_raw <- RCurl::getURL("https://cran.r-project.org/web/packages/dispRity/index.html")
version_CRAN <- strsplit(strsplit(version_raw, split = "\n<td>Version:</td>\n<td>")[[1]][2], split = "</td>\n</tr>\n<tr>\n<td>Depends:")[[1]][1]
version_CRAN <- 1.5
```

@@

Furthermore, don't forget to cite `R`:

> R Core Team (2020). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.

Bonus: you can also cite `ape` since the `dispRity` package heavily relies on it:

> Paradis E. & Schliep K. 2019. ape 5.0: an environment for modern phylogenetics and evolutionary analyses in R. Bioinformatics 35: 526-528.

@@

> Note that using the `rarefaction` argument also bootstraps the data. In these examples, the function bootstraps the data (without rarefaction) AND also bootstraps the data with the different rarefaction levels.

@@

Level | Name | Description | Source | 
------|------|---------------------------------------------------|--------|
2 | `ancestral.dist` | The distance between an element and its ancestor | `dispRity` |
2 | `angles` | The angle of main variation of each dimensions | `dispRity` |
2 | `centroids`<sup>1</sup> | The distance between each element and the centroid of the ordinated space | `dispRity` |
1 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln$area` |
1 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln$vol` |
2 | `deviations` | The minimal distance between each element and a hyperplane | `dispRity` |
1 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
2 | `displacements` | The ratio between the distance from a reference and the distance from the centroid | `dispRity` |
1 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | Donohue _et al._ (2013) |
1 | `func.div` | The functional divergence (the ratio of deviation from the centroid) | `dispRity` (similar to [`FD`](https://cran.r-project.org/web/packages/FD/index.html)`::dbFD$FDiv` but without abundance)|
1 | `func.eve` | The functional evenness (the minimal spanning tree distances evenness) | `dispRity` (similar to [`FD`](https://cran.r-project.org/web/packages/FD/index.html)`::dbFD$FEve` but without abundance)|
1 | `group.dist` | The distance between two groups | `dispRity` |
1 | `mode.val` | The modal value | `dispRity` |
1 | `n.ball.volume` | The hyper-spherical (*n*-ball) volume | `dispRity` |
2 | `neighbours` | The distance to specific neighbours (e.g. the nearest neighbours - by default) | `dispRity` |
2 | `pairwise.dist` | The pairwise distances between elements | [`vegan`](https://cran.r-project.org/web/packages/vegan/index.html)`::vegist` |
2 | `point.dist` | The distance between one group and the point of another group | `dispRity` |
2 | `quantiles` | The *n*th quantile range per axis | `dispRity` |
2 | `radius` | The radius of each dimensions | `dispRity` |
2 | `ranges` | The range of each dimension | `dispRity` |
2 | `span.tree.length` | The minimal spanning tree length | [`ape`](https://cran.r-project.org/web/packages/vegan/index.html) (but see also [`vegan`](https://cran.r-project.org/web/packages/vegan/index.html)`::spantree`) |
2 | `variances` | The variance of each dimension | `dispRity` |

@@

#### Between group metrics {#betweengroupmetricslist}

You can find detailed explanation on how between group metrics work [here](#betweengroupmetricsexplain).

##### `group.dist`

The `group.dist` metric allows to measure the distance between two groups in the multidimensional space.
This function needs to intake several groups and use the option `between.groups = TRUE` in the `dispRity` function.
It calculates the vector normal distance (euclidean) between two groups and returns 0 if that distance is negative.
Note that it is possible to set up which quantiles to consider for calculating the distances between groups.
For example, one might be interested in only considering the 95% CI for each group.
This can be done through the option `probs = c(0.025, 0.975)` that is passed to the `quantile` function.
It is also possible to use this function to measure the distance between the groups centroids by calculating the 50% quantile (`probs = c(0.5)`).

```{r}
## Creating a dispRity object with two groups
grouped_space <- custom.subsets(dummy_space,
                      group = list(c(1:5), c(6:10)))

## Measuring the minimum distance between both groups
summary(dispRity(grouped_space, metric = group.dist,
                 between.groups = TRUE))

## Measuring the centroid distance between both groups
summary(dispRity(grouped_space, metric = group.dist,
                 between.groups = TRUE, probs = 0.5))

## Measuring the distance between both group's 75% CI
summary(dispRity(grouped_space, metric = group.dist,
                 between.groups = TRUE, probs = c(0.25, 0.75)))
```

##### `point.dist`

The metric measures the distance between the elements in one group (`matrix`) and a point calculated from a second group (`matrix2`).
By default this point is the centroid but can be any point defined by a function passed to the `point` argument.
For example, the centroid of `matrix2` is the mean of each column of that matrix so `point = colMeans` (default).
This function also takes the `method` argument like previous one described above to measure either the `"euclidean"` (default) or the `"manhattan"` distances:

```{r}
## Measuring the distance between the elements of the first group
## and the centroid of the second group
summary(dispRity(grouped_space, metric = point.dist,
                 between.groups = TRUE))

## Measuring the distance between the elements of the second group
## and the centroid of the first group
summary(dispRity(grouped_space, metric = point.dist,
                 between.groups = list(c(2,1))))

## Measuring the distance between the elements of the first group
## a point defined as the standard deviation of each column
## in the second group
sd.point <- function(matrix2) {apply(matrix2, 2, sd)}
summary(dispRity(grouped_space, metric = point.dist,
                 point = sd.point, method = "manhattan",
                 between.groups = TRUE))
```
